---
output:
    md_document:
        variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)


readExample <- function(filename, from=NULL, to=NULL) {
    cont <- readLines(paste0('inst/examples/', filename))
    
    if (!is.null(from)) {
        from <- grep(paste0('^# ', from, '\\. '), cont)
        from <- from[1]
        from <- from + 2
        while (trimws(cont[from]) == '' && from < length(cont)) {
            from <- from + 1
        }
    } else {
        from <- 1
    }
    
    if (!is.null(to)) {
        to <- grep(paste0('^# ', to, '\\. '), cont)
        to <- to[1]
        to <- to - 2
        while (trimws(cont[to]) == '' && to > 1) {
            to <- to - 1
        }
    } else {
        to <- length(cont)
    }
    
    cont <- cont[seq.int(from, to)]
    
    cat('```\n')
    cat(cont, sep='\n')
    cat('```\n')
}
```


# crudtable

![crudtable build on travis-ci.org](https://travis-ci.org/beerda/crudtable.svg?branch=master)
![crudtable build on appveyor.com](https://ci.appveyor.com/api/projects/status/github/beerda/crudtable?branch=master&amp;svg=true)
![crudtable code coverage](https://codecov.io/gh/beerda/crudtable/branch/master/graph/badge.svg)
![crudtable in CRAN](http://www.r-pkg.org/badges/version/crudtable)

**crudtable** is an [R](https://www.r-project.org/) package that makes easy tabular data input in
[Shiny](https://shiny.rstudio.com/) web applications. With **crudtable**, the following operations
may be easily achieved:

* *CRUD* - **C**reate, **R**read, **U**pdate and **D**elete of data records in
  [DT](https://cran.r-project.org/web/packages/DT/index.html) DataTable and a modal edit dialog
  window;
* *validation* - ensuring the correct format of the user input;
* *database access* - storing the data into a database via the standardized [DBI](https://www.r-dbi.org/) package
  for [R](https://www.r-project.org/) or to a file.
  
  

## Getting Started

To install the latest development version from GitHub:

```
install.packages("remotes")
remotes::install_github("beerda/crudtable")
```


## A Minimal Working Example

A minimal Shiny app that uses **crudtable**:

```{r, results='asis'}
readExample('minimal/app.R')
```

First, a Data Access Object (DAO) is created with `dataFrameDao`.
DAO is a list structure that provides data access functions to the `crudTable` user interface.
In this example, a simple DAO is created that works with an in-memory data frame `CO2`. 
Alternatively, a SQL database may be connected with **crudtable**'s `sqlDao` DAO.

The UI part consists of `crudTableUI` that uses
[DT](https://cran.r-project.org/web/packages/DT/index.html)'s `DataTable` to view the dataset.
The **crudtable** UI also provides the *New record*, *Edit record* and *Delete record* buttons.

The server part consists of the call of the `crudTable` module that connects the `crudTableUI`
with the DAO.



## An Advanced Example

All the aspects and capabilities of the **crudtable** package will be shown in this advanced
example, which covers:

* access to an SQLite data table;
* custom input form user interface;
* validation of the user input;
* how to store values that are, rather than directly entered by the user, obtained programmatically.

First of all, let us import all the needed packages:

```{r, results='asis'}
readExample('advanced/app.R', NULL, 1)
```

We need `DBI` and `RSQLite` for database access, and `shinyjs` for JavaScript support. 

Next, we initialize the in-memory SQLite database engine and register the connection cleanup hook on
stop of Shiny. We also create an empty data frame `df` with columns: `date`, `service`,
`amount`, `discount`, `total` and `paid`. This data frame is saved into SQLite as table `'invoice'`.
We also create a Data Access Object (DAO) `dao` by calling the `sqlDao()` function:

```{r, results='asis'}
readExample('advanced/app.R', 1, 2)
```

For our convenience, we also create a constant list of service prices that will be used to populate
the select box with values:

```{r, results='asis'}
readExample('advanced/app.R', 2, 3)
```

We also want a custom edit dialog window with some pre-defined values and well defined ranges for
numeric inputs. We also add two read only input lines that will present some computed values to
the user. For that, we use the `disabled()` function of the `shinyjs` package. Note also the
namespacing of the input IDs by the `ns()` function, which is mandatory:

```{r, results='asis'}
readExample('advanced/app.R', 3, 4)
```

After the edit form UI is defined, we need to create the server part of the form handler. Since
we want to perform a lot of custom functionality, we code the server part in two steps. First, a
default form server handler is initialized by calling the `formServerFactory()` function:

```{r, results='asis'}
readExample('advanced/app.R', 4, 5)
```

`formServerFactory()` requires `dao` and a definition of `validators`. Validator is a mechanism
for restricting the input to certain criteria. If the user insert invalid input, an error message
is shown and the edit form dialog can not be submitted. In the piece of code above, we define two
types of validators: a custom validator bound to the `amount` data input, which tests the oddness
of the value. The second validator is `filledValidator` that ensures the data inputs are filled.
`filledValidator` is bound to all data inputs - we call `names(dao$getAttributes())` instead of
enumerating names of all data columns.

Now we can define our server-side handler of the edit form. First, `defaultFormServer` handler
must be called, which returns a list of useful reactive values and triggers. After that, we can
provide an observer that computes the read only inputs of the form. Note that we need to observe
the `res$loadTrigger()` here, which triggers everytime the data get loaded into the edit form.
This ensures that the computed values are initialized properly too. Note also that the server-side
handler must return the `res`, which is the result of `defaultFormServer`:

```{r, results='asis'}
readExample('advanced/app.R', 5, 6)
```

And that's nearly all. The last step is the initialization of the Shiny app. We use `crudTableUI`
on the client side and we call the `crudTable` module on the server side. The latter gets `dao`,
`myFormUI` and `myFormServer` as arguments:

```{r, results='asis'}
readExample('advanced/app.R', 6, NULL)
```

Note that it is not needed to call the `useShinyjs()` function in the UI of the Shiny application
since the **crudtable** package does it internally by itself.

The complete advanced example is as follows:

```{r, results='asis'}
readExample('advanced/app.R')
```
